MSstatsPTM: Statistical relative quantification of post-translational modifications in bottom-up mass spectrometry-based proteomics

### 2022-06-20 main/MSstatsPTM_Main.tex
general rewording and updated table

Additional details on every dataset are available in Supplementary Section 2.

@@

\medskip \noindent {\bf Evaluation:} We evaluated the ability of the statistical methods to correctly detect which PTMs were simulated with changes between conditions and which were not. We evaluated the methods ability to avoid false positives (i.e. specificity), accurately estimate the fold change between conditions, and analyzed the sensitivity of detecting differentially abundant PTMs. Each method was used to analyze changes in PTM abundance in the presence of confounding with changes in the unmodified protein, and with this confounding removed.

@@

\medskip \noindent {\bf Evaluation:} The methods were evaluated in the same way as above. We evaluated their ability to correctly detect PTM's specificity, fold change estimation, and sensitivity. These statistics were analyzed both in the presence of, and without, confounding with changes in the unmodified protein.

@@

\medskip \noindent {\bf Limma} Modifications of $t$-test such as moderated $t$-test with Limma were also proposed~\cite{Ritchie_15a}\cite{Zhu}\cite{Chappell:2021}. The inputs to the Limma based methods are the same as the inputs to $t$-test. The intensities of individual features from modified peptides, $y_{ijk}^{\ast}$, can be used directly (in the presence of confounding with the unmodified protein). To remove confounding we transform the input by taking the intensity ratios of modified and unmodified peptide features, $\hat{u}^{\ast}_{ij}$, calculated in the same way as Equation \ref{eq:adj_estimation}. Limma then fits a linear model using the transformed inputs to estimate $\hat{\Delta}$. To estimate $\widehat{SE(\hat{\Delta})}$, Limma leverages empirical bayes moderation~\cite{Robbins1992} to moderate the model variance of the individual PTM using the shared global variance of all PTMs modeled in the experiment. For a single PTM $m$ with an estimated fold change $\hat{\Delta}_m$, the estimated variance is calculated as $s^{adj}_{m} = P(s_{m}| s_\Theta)$ where $s_{m}$ is the variance resulting from the linear model of PTM $m$ and $s_\Theta$ is the global variance across all PTMs. The statistical significance is determined using $\hat{\Delta}_m$ and $\widehat{SE(\hat{\Delta})}$ calculated using the adjusted variance $s^{adj}_{m}$.

@@

Limma provides an alternative that overcomes some of the limitations of $t$-test. First, utilizing empirical bayes moderation can increase the sensitivity of the analysis and is especially useful in cases of low replicates. Additionally, using a linear model allows for the sharing of variance information across comparisons of interest. This means we can more accurately measure the variance as the number of comparisons increases.

@@

$MSstats$~\cite{Choi:2014} is designed for protein significance analysis and statistical relative quantification of proteins and peptides in global, targeted and data-independent proteomics. The package takes as input a list of intensities of individual features, denoted by $y_{ijk}$, used to characterize the identified protein. For each protein, the feature log-intensities are first summarized into a single value per protein per run in the subplot model. This is done using Tukey's median polish (TMP)~\cite{Tukey:1977}, a simple and robust procedure, which iteratively fits a two-way additive model with the effects of run and feature, summarizing them into a single value per run. These summarized values are then used as input to fit a linear model based on the family of linear mixed-effects models \cite{Bolker2009} \cite{Faraway:2006}. These methods are implemented in the packages $MSstats$ and $MSstatsTMT$~\cite{Huang:2020} and include functionalities for end to end quantification of proteins and differential analysis between experimental conditions.

The $MSstats$ analysis pipeline has a number advantages over simpler methods. First, the approach extends to many types of complex experimental design, including unbalanced designs, and additional sources of variation, such as fractionation. The methods include optional implementation of empirical bayes moderation, described previously, when applied to TMT experiments~\cite{Huang:2020}. Additionally, the summarization step includes options for missing value imputation, using an accelerated failure time (AFT) model~\cite{Wei:1992}, and has been shown to be more robust compared to using individual features~\cite{Choi:2014}.

@@

We propose a statistical method for detecting changes in PTMs that allows us to remove confounding with changes in the global protein and is applicable to experiments with complex designs.   To do this we leverage the existing methods in $MSstats$ and extend them to remove confounding between changes in the PTM and global protein. The method takes as input the spectral features of the PTM, $y_{ijk}^{\ast}$, and it's corresponding global protein, $y_{ijk}$. The features are summarized separately using the methods described previously for $MSstats$, resulting in run level values $\hat{y}_{1++}^{\ast}$ for the modified peptide and $\hat{y}_{1++}$ for the unmodified peptide. These summarized values are then used as the input to our models. Two separate linear mixed-effects models are fit, one using the summarized PTM values and another using the summarized global protein values. The models give estimates of the difference in log-abundance of the underlying PTM, $\hat{\Delta}_{PTM}$, and global protein, $\hat{\Delta}_{protein}$. Finally, the adjusted difference, $\hat{\Delta}$, is given by the difference in log-abundance of the PTM site, subtracted by the difference in log-abundance of the underlying protein.

@@

\todo{DK: I'm not sure where this part should be. could move it to the first paragraph of this section, or in the extensions to complex design}
Finally, this approach allows us to model experiments with separate designs for the PTM and global profiling runs. This includes differing number of runs, biological replicates, or types of acquisition methods. This is achieved through fully modeling the PTM and global protein separately before adjusting. Because we are not combining on the replicate level, the number of replicates for each group do not need to match for adjustment. This makes the proposed method more flexible and applicable to a wider variety of experimental designs.

@@

The performance of the proposed method was evaluated on simulated and spike-in datasets with known ground truth, as well as biological experimental data where the ground truth was not known. For the experiments where ground truth was known, we calculated the number of true positives (TP), false positives (FP), true negatives (TN), and false negatives (FN). True positives were modified peptides that were differential after accounting for changes in global protein abundance. True negatives were modified peptides that were not differential after accounting for changes in global protein abundance. To determine positives and negatives we adjusted the p-values for multiple testing and used an $\alpha=.05$ cutoff. These results were then compared to the ground truth to determine TP and TN. We used these results to calculate the false discovery rate (FDR = FP/(TP+FP)), recall rate (TP/(TP + FN), and accuracy ((TP + TN) / (TP + TN + FP + FN)). Finally, the summary statistics were compared across methods to analyze method performance.

@@

The proposed approach was evaluated against two-sample $t$-test and Limma in computer simulations. 
In Figure \ref{fig:sim1_fdr} we see that not accounting for changes in the unmodified protein resulted in a high false discovery rate. After adjusting for changes in the unmodified protein, all methods performed similarly in terms of FDR. In a clean simulation, recall and accuracy of the proposed approach and Limma performed similarly, although Limma performed slightly better in simulations with fewer biological replicates. In contrast, the performance of the $t$-test method lagged behind the others. Two-sample $t$-test only uses data within the groups of interest while ignoring the remainder of the data, leading to similar performance across all simulations. In contrast, the proposed approach and Limma leverage all available information, which resulted in improved power in simulations with more conditions.

@@

In this simulation the proposed method outperformed $t$-test and Limma. Changes in unmodified protein level still needed to be accounted for to control the FDR. Once controlling for changes in the unmodified protein, the proposed method outperformed the other methods, as seen in Figure \ref{fig:sim2_acc}. The proposed method calibrated model accuracy and recall rate well, even when the number of replicates were low. Additionally, when comparing the fold change estimation across all modified peptides, the proposed method showed a tighter distribution of estimated fold changes around the true fold change, as compared to $t$-test and Limma. Specifically, the inner quartile range (IQR) of the estimated fold change for the proposed method was on average 21.8\% smaller than Limma's IQR and 10.4\% smaller than $t$-test's IQR. This fold change comparison can be seen in Supplementary Figure \ref{supp-fig:fc_boxplot}.

@@

In Figure \ref{fig:spikein_prop_volcano} we can clearly see the fold change of the red labeled spike-in peptides was only correctly estimated when accounting for changes in the unmodified protein abundance. Additionally, the background peptides, serving as the null model, show many false positives before adjustment. After adjustment the number of false positives decreased significantly. Specifically, for the proposed method, the number of false positives went from 20.88\% to 1.84\% after adjustment was applied. While the proposed method and Limma both correctly estimated the fold change of the spike-in peptides, Limma resulted in the majority having an insignificant adjusted p-value. In this case, using Limma would have resulted in missing the majority of differentially modified peptides.

@@

The results of this experiment are summarized in Figure \ref{fig:data4_plots}. In Figure \ref{fig:data4_venn_diagram} the number of significant modified peptides before and after adjustment is shown. We can see that more peptides became insignificant after adjustment than became significant. 3,236 modified peptides became insignificant, 1,352 became significant, while 4,282 were significant in both models. For the peptides that became insignificant in the adjusted model, their change in abundance was mainly due to changes in global protein abundance. In contrast, for peptides that became significant after adjustment, their true abundance change was masked by underlying changes in the unmodified protein. Both of these issues were corrected in adjustment, and the deconfounded abundance change is shown. An additional question that must be addressed is if the decrease in significant peptides is due to the increased variance that comes from adjustment. This was tested by looking for modified peptides whose adjusted log fold change was within 10\% of the unadjusted log fold change but became insignificant after adjustment. In other words, the fold change was the same between models but variance increased. When this test was applied on this experiment, only one peptide became insignificant due to an increase in variance. Thus we can conclude that the drop off in significant peptides was due to changes in global protein abundance.

@@

The results of this experiment are summarized in Figure \ref{fig:data5_plots}. In Figure \ref{fig:data5_venn_diagram} the number of significant modified peptides before and after adjustment is shown. Again more PTMs became insignificant after adjustment than became significant. 19,286 peptides became insignificant, 4,947 became significant, and 41,552 were significant in both models. Again we tested if the decrease in significant peptides was due to increased variance or if it was due to removing convolution with changes in the umodified protein. This was tested by looking for modified peptides whose adjusted log fold change was within 10\% of the unadjusted log fold change but became insignificant after adjustment. When this test was applied, 548 peptides became insignificant due to an increase in variance. This is only 3.4\% of the total peptides that became insignificant after adjustment. Thus we can conclude that the drop off in significant peptides was mainly due to changes in global protein abundance.

@@

As discussed in Section \ref{sec:exp_proc_dataset6}, there was no unmodified global profiling run performed in this experiment. Once identification and quantification of the Ubiquitination profiling was performed, peptides which were unmodified were extracted and used in place of a global profiling run. This resulted in a lack of overlap between modified and unmodified peptides. Any modified peptide without a corresponding unmodified protein could not be adjusted. Of the 10,799 ubiquitin sites identified, only 4,526 had a corresponding unmodified protein and could be adjusted. Additionally, the lack of a separate global profiling run resulted in low feature counts for the unmodified protein model.

The results of this experiment are summarized in Figure \ref{fig:data6_plots}. After adjusting for changes in the unmodified protein, there were fewer significant modified peptides. In total, 2709 modified peptides became insignificant, while only 547 became significant. However, this was mainly due to not having a global profiling run, resulting in a lack of overlap between modified and unmodified peptides. In the bottom plot, only modified peptides that could be adjusted are shown. Here there were much fewer peptides that became insignificant after adjustment. 726 modified peptides became insignificant, 547 became significant, and 1,078 were significant in both models. As in the previous experiments, we can check if these peptides became insignificant due to an increase in variance. To check this we looked at modified peptides whose adjusted fold change was within 10\% of their unadjusted fold change and which became insignificant after adjustment. In this experiment there were only 25 modified peptides that meet this criteria.

@@

Our results show that the proposed approach for modeling and summarization leads to more sensitive PTM significance analysis and more accurate and precise quantification. The gain is due to a more efficient use of the data, and to a more accurate understanding of the systematic and random variations. The proposed framework can be extended beyond the experimental designs with variation from multiple sources discussed above. Although demonstrated here on DDA, is also applicable to DIA, SRP and PRM acquisitions. Additionally, the approach can handle experiments with modified peptides processed using label-free methods and unmodified peptides processed using TMT labeling, or vice versa. In this case summarization and modeling is still done separately for both the modified and unmodified data, and then combined after modeling.

@@

Additionally, abundance levels of PTM sites can be convoluted with each other if there are multiple modification sites per peptide, or convoluted with changes in the unmodified peptide (as opposed to the unmodified protein). In the current implementation the effect of a specific modification in a peptide with multiple modifications cannot be quantified. One potential solution to this is to measure the abundance of peptides with one modification and use this to adjust the peptide with multiple sites to remove the convolution. However, this method would likely run into challenges due to sparsity of features for modified peptides with both a single and multiple modification sites. A more complex approach to addressing this problem is most likely necessary.

@@

%%%% Summary of controlled mixtures
\begin{table}[ht]
\centering
\caption*{Experimental Datasets Overview}
\begin{tiny}
\begin{tabular}{|c|c|ccccc|cc|}
\hline
%%%% colnames of table
& \multirow{2}{*}{Dataset} & No. of & No. of bio. & No. of & No. of mod. & No. of unmod.&  Data & \multirow{2}{*}{Analysis}\\
&  & Conditions & replicates & proteins & features/site & peptides/protein & availability & \\
\hline
\hline
 &&&&&&&& \\[-0.05in]
%%%% DDA:ControlledMix
\multirow{5}{*}{\rotatebox[origin=c]{90}{Known}}  \multirow{5}{*}{\rotatebox[origin=c]{90}{Ground}} \multirow{5}{*}{\rotatebox[origin=c]{90}{Truth}} & Computer simulation 1 - Label-free & 2/3/4 & 2/3/5/10 & 1000 & 10 & 10 & \multicolumn{2}{c|}{\href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github}} \\
 &&&&&&&& \\%[-0.05in]
%\hline 
%&&&&&&&& \\[-0.05in]
% \cline{2-9}
% &&&&&&&& \\[-0.05in]
%%% DDA iPRG2015
& Computer simulation 2 - Missing and low features & 2/3/4 & 2/3/5/10 & 1000 & 2 & 10 & \multicolumn{2}{c|}{\href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github}} \\
 &&&&&&&& \\%[0.05in]
%\hline 
% &&&&&& \\[-0.05in]
%%% DDA Cox2014
& SpikeIn benchmark - Ubiquitination - Label-free& 4 & 2 & 3198 & 1.37 & 10.17 & \href{https://massive.ucsd.edu/ProteoSAFe/private-dataset.jsp?task=c4c583ecf7f941cdac87f7a4f872517b}{MSV000088971} & \href{https://massive.ucsd.edu/ProteoSAFe/private-dataset.jsp?task=c4c583ecf7f941cdac87f7a4f872517b}{TBD} \\
 &&&&&&&& \\%[-0.05in]
\hline
\multicolumn{6}{c}{ } \\ [0.02in]
\hline 
 &&&&&&&& \\%[-0.05in]
%%%%%%%%%%%%%%%%%%
%%%% biological study
%%%%%%%%%%%%%%%%%%
%% DDA Meierhofer
\multirow{3}{*}{\rotatebox[origin=c]{90}{Biological}} \multirow{3}{*}{\rotatebox[origin=c]{90}{Experiment}} & Human - Ubiquitination - 1mix-TMT & 6 & 2 & 3550 & 1.21 & 10.75 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=b6f0c74c234247678fb0888c6df1f225}{MSV000088966} & \href{https://massive.ucsd.edu/ProteoSAFe/reanalysis_container.jsp?task=b639fa5d0ff24ce78abdca13f4b18c0a}{RMSV000000356}\\
&&&&&&&& \\%[-0.05in]
% \cline{2-9}
% &&&&&&& \\[-0.05in]
%%%% SRM Picotti 2009
& Mouse - Phosphorylation - 2mix-TMT& 6 & 4 & 5002 & 1.67 & 10.95 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=4878d777c6b34cf8aaf8477e93140c4d}{MSV000085565} & \href{https://massive.ucsd.edu/ProteoSAFe/reanalysis_container.jsp?task=79406d32c1f9413daf7756e6eeb24968}{RMSV000000357}\\
&&&&&&&& \\%[-0.05in]
% \cline{2-9}
% &&&&&&& \\[-0.05in]
%%%% SRM Surinova 2015, training
& Human - Ubiquitination - Label-free & 4 & 2 & 3863 & 1.40 & 1.65 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=1b516164de5345108b40b75147dd58b5}{MSV000078977} & \href{https://massive.ucsd.edu/ProteoSAFe/reanalysis_container.jsp?task=f9e6117e00794ce5b2e8802601be16bc}{RMSV000000358}\\ [0.02in]
 &&&&&&&& \\
 \hline
\end{tabular}\\
\end{tiny}
\caption{ \small {\bf Simulated and experimental datasets.} 
``Dataset'' is the dataset code name.
``Data availability'' shows the ID of the MassIVE.quant repository or the GitHub repository. ``Analysis'' shows the ID of the MassIVE.quant reanalysis container, containing analysis code and modeling results. All the experiments were conducted in data-dependent acquisition (DDA) mode. \todo{Could you add to the table the number of proteins, and the average number of modified/unmodified peptides per protein} \todo{If the figures are at the end, I think the table should also be at the end}
\todo{Could you check the table and the entire section: when we say 'number of biological replicates', we typically mean per condition. But it seems that some descriptions imply the total number of biological samples across conditions. Could you clarify that we use number of biological replicates per condition throughout? (E.g, this helps calculate the degrees of freedom in balanced designs)}
}
\label{tab:dataDescription}
\end{table}

>>>>

Additional details on all experiments are available in Supplementary Section 2.

@@

\medskip \noindent {\bf Evaluation:} We evaluated the ability of the statistical methods to correctly detect which PTMs were simulated with changes between conditions and which were not. We gauged the methods ability to avoid false positives (i.e. specificity), accurately estimate the fold change between conditions, and analyzed the sensitivity of detecting differentially abundant PTMs. The evaluation was performed both in the presence of convolution with changes in the unmodified protein and after applying adjustment to correct for convolution.

@@

\medskip \noindent {\bf Evaluation:} The methods were evaluated as above. We evaluated their ability to correctly detect PTM's specificity, fold change estimation, and sensitivity. These statistics were analyzed both in the presence of, and without, convolution with changes in the unmodified protein.

@@

\medskip \noindent {\bf Limma} Modifications of $t$-test such as moderated $t$-test with Limma were also proposed~\cite{Ritchie_15a}\cite{Zhu}\cite{Chappell:2021}. The inputs to  Limma are the same as the inputs to $t$-test. The intensities of individual features from modified peptides, $y_{ijk}^{\ast}$, can be used directly (in the presence of confounding with the unmodified protein). To remove confounding we transform the input by taking the intensity ratios of modified and unmodified peptide features, $\hat{u}^{\ast}_{ij}$, calculated in the same way as Equation \ref{eq:adj_estimation}. Limma then fits a linear model using the transformed inputs to estimate $\hat{\Delta}$. To estimate $\widehat{SE(\hat{\Delta})}$, Limma leverages empirical bayes moderation~\cite{Robbins1992} to moderate the model variance of the individual PTM using the shared global variance of all PTMs modeled in the experiment. For a single PTM $m$ with an estimated fold change $\hat{\Delta}_m$, the estimated variance is calculated as $s^{adj}_{m} = P(s_{m}| s_\Theta)$ where $s_{m}$ is the variance resulting from the linear model of PTM $m$ and $s_\Theta$ is the global variance across all PTMs. The statistical significance is determined using $\hat{\Delta}_m$ and $\widehat{SE(\hat{\Delta})}$ calculated using the adjusted variance $s^{adj}_{m}$.

@@

Limma provides an alternative that overcomes limitations of $t$-test. First, utilizing empirical bayes moderation can increase the sensitivity of the analysis and is especially useful in cases of low replicates. Additionally, using a linear model allows for the sharing of variance information across comparisons of interest. This means we can more accurately measure the variance as the number of comparisons increases.

@@

\todo{DK: This section could use some rewording}

$MSstats$~\cite{Choi:2014} is designed for protein significance analysis and statistical relative quantification of proteins and peptides in global, targeted and data-independent proteomics. The package takes as input a list of intensities of individual features, denoted by $y_{ijk}$, used to characterize the identified protein. For each protein, the feature log-intensities are first summarized into a single value per protein per run in the subplot model. This is done using Tukey's median polish (TMP)~\cite{Tukey:1977}, a simple and robust procedure, which iteratively fits a two-way additive model with the effects of run and feature, summarizing them into a single value per run. These summarized values are then used as input to fit a linear model based on the family of linear mixed-effects models \cite{Bolker2009} \cite{Faraway:2006}. The output of this model is the quantity of interest $\hat{\Delta}$ and the standard error, $\widehat{SE(\hat{\Delta})}$. These methods are implemented in the packages $MSstats$ and $MSstatsTMT$~\cite{Huang:2020} and include functionalities for end to end quantification of proteins and differential analysis between experimental conditions.

The $MSstats$ analysis pipeline has a number advantages over simpler methods. First, the approach extends to many types of complex experimental design, including unbalanced designs and multiple sources of variation, such as fractionation. The methods include optional implementation of empirical bayes moderation, described previously, when applied to TMT experiments~\cite{Huang:2020}. Additionally, the summarization step includes options for missing value imputation, using an accelerated failure time (AFT) model~\cite{Wei:1992}, and has been shown to be more robust compared to using individual features~\cite{Choi:2014}.

@@

We propose a statistical method for detecting changes in PTMs that allows us to remove confounding with changes in the global protein and is applicable to experiments with complex designs.   To do this, we leverage the existing methods in $MSstats$ and extend them to remove confounding between changes in the PTM and global protein. The method takes as input the spectral features of the PTM, $y_{ijk}^{\ast}$, and it's corresponding global protein, $y_{ijk}$. The features are summarized separately using the methods described previously for $MSstats$, resulting in run level values for the PTM, $\hat{y}_{1++}^{\ast}$, and unmodified peptide, $\hat{y}_{1++}$. These summarized values are then used as the input to our models. Two separate linear mixed-effects models are fit, one using the summarized PTM values and another using the summarized global protein values. The models give estimates of the difference in log-abundance of the underlying PTM, $\hat{\Delta}_{PTM}$, and global protein, $\hat{\Delta}_{protein}$. Finally, the adjusted difference, $\hat{\Delta}$, is given by the difference in log-abundance of the PTM site, subtracted by the difference in log-abundance of the underlying protein.

@@

\todo{DK: I'm not sure where this part should be. could move it to the first paragraph of this section, or in the extensions to complex design}
This approach allows us to model experiments with separate designs for the PTM and global profiling runs. This includes differing number of runs, biological replicates, and/or types of acquisition methods. This is achieved through fully modeling the PTM and global protein separately before adjusting. Because we are not combining on the replicate level, the number of replicates for each group do not need to match for adjustment. This makes the proposed method more flexible and applicable to a wider variety of experimental designs.

@@

The performance of the proposed method was evaluated on simulated and spike-in datasets with known ground truth, as well as biological experimental data where the ground truth was not known. When ground truth was known, we calculated the number of true positives (TP), false positives (FP), true negatives (TN), and false negatives (FN). True positives were PTMs that were expected to be differential after accounting for changes in global protein abundance. True negatives were PTMs that were not expected to be differential after accounting for changes in global protein abundance. To determine positives and negatives we adjusted the p-values for multiple testing and used an $\alpha=.05$ cutoff. These results were then compared to the ground truth to determine TP and TN. Summary statistics were calculated, including the false discovery rate (FDR) (FP/(TP+FP)), recall rate (TP/(TP + FN), and accuracy ((TP + TN) / (TP + TN + FP + FN)). Finally, the summary statistics were compared across methods to analyze method performance.

@@

The proposed approach was evaluated against two-sample $t$-test and Limma in computer simulations. 
In Figure \ref{fig:sim1_fdr} we see that not accounting for changes in the unmodified protein resulted in a high FDR. After adjusting for changes in the unmodified protein, all methods performed similarly in terms of FDR. In a clean simulation, recall and accuracy of the proposed approach and Limma performed similarly, although Limma performed slightly better in simulations with fewer biological replicates. In contrast, the performance of the $t$-test method lagged behind the others. Two-sample $t$-test only uses data within the groups of interest while ignoring the remainder of the data, leading to similar performance across all simulations. In contrast, the proposed approach and Limma leverage information across comparisons, resulting in improved power in simulations with more comparisons.

@@

In simulations with missing values and low features the proposed method outperformed $t$-test and Limma. Changes in unmodified protein level still needed to be accounted for to control the FDR. Once controlling for changes in the unmodified protein, the proposed method outperformed the other methods, as seen in Figure \ref{fig:sim2_acc}. The proposed method calibrated model accuracy and recall rate well, even when the number of replicates were low. Additionally, when comparing the fold change estimation across all modified peptides, the proposed method showed a tighter distribution of estimated fold changes around the true fold change, as compared to $t$-test and Limma. Specifically, the inner quartile range (IQR) of the estimated fold change for the proposed method was on average 21.8\% smaller than Limma's IQR and 10.4\% smaller than $t$-test's IQR. This fold change comparison can be seen in Supplementary Figure \ref{supp-fig:fc_boxplot}.

@@

In Figure \ref{fig:spikein_prop_volcano} we can clearly see the fold change of the red labeled spike-in peptides was only correctly estimated when accounting for changes in the unmodified protein abundance. Additionally, the background peptides, serving as the null model, show many false positives before adjustment. After adjustment the number of false positives decreased significantly. Specifically, for the proposed method, the number of false positives went from 20.88\% to 1.84\% after adjustment was applied. While the proposed method and Limma both correctly estimated the fold change of the spike-in peptides, using Limma resulted in many insignificant adjusted p-values. Using Limma would have resulted in missing the majority of differential PTMs.

@@

The results of this experiment are summarized in Figure \ref{fig:data4_plots}. Figure \ref{fig:data4_venn_diagram} shows the number of significant modified peptides before and after adjustment. More peptides became insignificant after adjustment than became significant. 3,236 modified peptides became insignificant, 1,352 became significant, while 4,282 were significant in both models. For the peptides that became insignificant in the adjusted model, their change in abundance was mainly due to changes in global protein abundance. In contrast, for peptides that became significant after adjustment, their true abundance change was masked by underlying changes in the unmodified protein. Both of these issues were corrected in adjustment, and the unconvoluted abundance change is shown. An additional question that must be addressed is if the decrease in significant peptides is due to the increased variance that comes from adjustment. This was tested by looking for modified peptides whose adjusted log fold change was within 10\% of the unadjusted log fold change but became insignificant after adjustment. In other words, the fold change was the same between models but variance increased. When this test was applied on this experiment, only one peptide became insignificant due to an increase in variance. Thus we can conclude that the drop off in significant peptides was due to changes in global protein abundance.

@@

The results of this experiment are summarized in Figure \ref{fig:data5_plots}. Figure \ref{fig:data5_venn_diagram} shows the number of significant modified peptides before and after adjustment. Again more PTMs became insignificant after adjustment than became significant. 19,286 peptides became insignificant, 4,947 became significant, and 41,552 were significant in both models. Again we tested if the decrease in significant peptides was due to removing convolution with changes in the umodified protein or increased variance after adjustment was applied. This was tested by looking for modified peptides whose adjusted log fold change was within 10\% of the unadjusted log fold change but became insignificant after adjustment. When this test was applied, 548 peptides became insignificant due to an increase in variance. This is only 3.4\% of the total peptides that became insignificant after adjustment. Thus we can conclude that the drop off in significant peptides was mainly due to changes in global protein abundance.

@@

As discussed in Section \ref{sec:exp_proc_dataset6}, there was no unmodified global profiling run performed in this experiment. Once identification and quantification of the Ubiquitination profiling was performed, peptides which were unmodified were extracted and used in place of a global profiling run. This resulted in a lack of overlap between modified and unmodified peptides. Any modified peptide without a corresponding unmodified protein could not be adjusted. Of the 10,799 ubiquitin sites identified, only 4,526 had a corresponding unmodified protein and could be adjusted. Additionally, the lack of a separate global profiling run resulted in low feature counts for the unmodified protein model compared to other experiments, seen in Table \ref{tab:dataDescription}.

The results of this experiment are summarized in Figure \ref{fig:data6_plots}. After adjusting for changes in the unmodified protein, there were fewer significant modified peptides. In total, 2709 modified peptides became insignificant, while only 547 became significant. However, this was mainly due to not having a global profiling run, resulting in a lack of overlap between modified and unmodified peptides. In the bottom plot, only modified peptides that could be adjusted are shown. Here there were much fewer peptides that became insignificant after adjustment. 726 modified peptides became insignificant, 547 became significant, and 1,078 were significant in both models. Only 25 PTMs became insignificant due to increased variance from adjustment.

@@

Our results show that the proposed approach for modeling and summarization leads to more sensitive PTM significance analysis and more accurate and precise quantification. The gain is due to a more efficient use of the data, and to a more accurate understanding of the systematic and random variations. The proposed framework can be extended beyond the experimental designs with variation from multiple sources discussed above. Although demonstrated here on DDA, is also applicable to DIA, SRM and PRM acquisitions. Additionally, the approach can handle experiments with modified peptides processed using label-free methods and unmodified peptides processed using TMT labeling, or vice versa. In this case summarization and modeling is still done separately for both the modified and unmodified data, and then combined after modeling.

@@

Additionally, abundance levels of PTM sites can be convoluted with each other if there are multiple modification sites per peptide, or convoluted with changes in the unmodified peptide (as opposed to the unmodified protein). In the current implementation the effect of a specific modification in a peptide with multiple modifications cannot be quantified. One potential solution to this is to measure the abundance of peptides with one modification and use this to adjust the peptide with multiple sites to remove the convolution. However, this method would likely run into challenges due to sparsity of features for modified peptides with both a single and multiple modification sites. A more complex approach to addressing this problem is likely necessary.

@@

%%%% Summary of controlled mixtures
\begin{table}[ht]
\centering
\caption*{Experimental Datasets Overview}
\begin{tiny}
\begin{tabular}{|c|c|ccccc|cc|}
\hline
%%%% colnames of table
& \multirow{2}{*}{Dataset} & No. of & No. of bio. & No. of mod. & No. of mod. & No. of unmod.&  Data & \multirow{2}{*}{Analysis}\\
&  & Conditions & replicates & peptides & features/site & peptides/protein & availability & \\
\hline
\hline
 &&&&&&&& \\[-0.05in]
%%%% DDA:ControlledMix
\multirow{5}{*}{\rotatebox[origin=c]{90}{Known}}  \multirow{5}{*}{\rotatebox[origin=c]{90}{Ground}} \multirow{5}{*}{\rotatebox[origin=c]{90}{Truth}} & Computer simulation 1 - Label-free & 2/3/4 & 2/3/5/10 & 1,000 & 10 & 10 & \multicolumn{2}{c|}{\href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github}} \\
 &&&&&&&& \\%[-0.05in]
%\hline 
%&&&&&&&& \\[-0.05in]
% \cline{2-9}
% &&&&&&&& \\[-0.05in]
%%% DDA iPRG2015
& Computer simulation 2 - Missing and low features & 2/3/4 & 2/3/5/10 & 1,000 & 2 & 10 & \multicolumn{2}{c|}{\href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github}} \\
 &&&&&&&& \\%[0.05in]
%\hline 
% &&&&&& \\[-0.05in]
%%% DDA Cox2014
& SpikeIn benchmark - Ubiquitination - Label-free& 4 & 2 & 12,137 & 1.37 & 10.17 & \href{https://massive.ucsd.edu/ProteoSAFe/private-dataset.jsp?task=c4c583ecf7f941cdac87f7a4f872517b}{MSV000088971} & \href{https://massive.ucsd.edu/ProteoSAFe/private-dataset.jsp?task=c4c583ecf7f941cdac87f7a4f872517b}{TBD} \\
 &&&&&&&& \\%[-0.05in]
\hline
\multicolumn{6}{c}{ } \\ [0.02in]
\hline 
 &&&&&&&& \\%[-0.05in]
%%%%%%%%%%%%%%%%%%
%%%% biological study
%%%%%%%%%%%%%%%%%%
%% DDA Meierhofer
\multirow{3}{*}{\rotatebox[origin=c]{90}{Biological}} \multirow{3}{*}{\rotatebox[origin=c]{90}{Experiment}} & Human - Ubiquitination - 1mix-TMT & 6 & 2/1 & 8,848 & 1.21 & 10.75 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=b6f0c74c234247678fb0888c6df1f225}{MSV000088966} & \href{https://massive.ucsd.edu/ProteoSAFe/reanalysis_container.jsp?task=b639fa5d0ff24ce78abdca13f4b18c0a}{RMSV000000356}\\
&&&&&&&& \\%[-0.05in]
% \cline{2-9}
% &&&&&&& \\[-0.05in]
%%%% SRM Picotti 2009
& Mouse - Phosphorylation - 2mix-TMT& 6 & 4/3 & 26,433 & 1.67 & 10.95 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=4878d777c6b34cf8aaf8477e93140c4d}{MSV000085565} & \href{https://massive.ucsd.edu/ProteoSAFe/reanalysis_container.jsp?task=79406d32c1f9413daf7756e6eeb24968}{RMSV000000357}\\
&&&&&&&& \\%[-0.05in]
% \cline{2-9}
% &&&&&&& \\[-0.05in]
%%%% SRM Surinova 2015, training
& Human - Ubiquitination - Label-free & 4 & 2 & 10,799 & 1.40 & 1.65 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=1b516164de5345108b40b75147dd58b5}{MSV000078977} & \href{https://massive.ucsd.edu/ProteoSAFe/reanalysis_container.jsp?task=f9e6117e00794ce5b2e8802601be16bc}{RMSV000000358}\\ [0.02in]
 &&&&&&&& \\
 \hline
\end{tabular}\\
\end{tiny}
\caption{ \small {\bf Simulated and experimental datasets.} 
``Dataset'' is the dataset code name. ``No. of bio. replicates'' shows the number of biological replicates per condition. Simulations were generated with different numbers of replicates. The design of two Biological Experiments were unbalanced with unequal replicates per condition. ``No. of mod. features/site'' is the number of features used to estimate the abundance of a single modification. A feature is a peptide and its corresponding charge. ``No. of unmod. peptides/protein'' is the number of peptides without modifications that were used to estimate the global protein abundance. ``Data availability'' shows the ID of the MassIVE.quant repository or the GitHub repository. ``Analysis'' shows the ID of the MassIVE.quant reanalysis container, containing analysis code and modeling results. All the experiments were conducted in data-dependent acquisition (DDA) mode. }

\label{tab:dataDescription}
\end{table}
